Austral Spring Extratropical Southern Hemisphere zonally assymetric circulation \thanks{The research was supported by UBACyT20020170100428BA and the CLIMAX Project funded by Belmont Forum/ANR-15-JCL/-0002-01. Elio Campitelli was supported by a PhD grant from CONICET, Argentina.}

### 2022-03-25 analysis/paper/shceof.tex
Render

The Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope (Irving and Simmonds 2015).
The latter makes possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
With this approach, Irving and Simmonds (2015) showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.

@@

An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions (Horel 1984).
This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain on water level to study tide-river interactions (Laurel-Castillo and Valle-Levinson 2020) and temperature and precipitation anomalies (e.g. Krokhin and Luxemburg 2007; Gelbrecht, Boers, and Jürgen Kurths 2018).
To our knowledge, complex EOF has not been applied in the spatial domain to capture the phase-varying nature of planetary waves.

@@

In Section 2 we describe the methods.
In Section 3.1, we analyse the spatial patterns of each complex EOF.
In Section 3.2 we study the spatial regressions with geopotential height and temperature anomalies.
In Section 3.3, we analyse the relationship between cEOF1 and Ozone, and between cEOF2 with the SAM and the PSA.
In Section 3.4 we study tropical forcings that explain the variability of each cEOF.
Finally, in Section 3.5 we show the relationship between these modes of variability and precipitation anomalies in South America and Oceania.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree~longitude by 2.5\degree~latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; Hersbach et al. (2019){]} for the period 1979-2019.
Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 (Bell et al. 2020) to describe long-term trends.
We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK (Adams, Swartztrauber, and Sweet 1999) and we computed horizontal wave activity fluxes following Plumb (1985).
Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 (Huang et al. 2017) and precipitation monthly data from the CPC Merged Analysis of Precipitation (P. Xie and Arkin 1997), with a 2.5\degree~resolution in latitude and longitude.
The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

@@

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period according (Baldwin and Thompson 2009).
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022).
Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated -- in agreement with Mo and Paegle (2001) -- as the third and fourth leading EOF of seasonal mean anomalies for SH 500 hPa geopotential height.

@@

In standard way of applying EOF analysis zonal waves may appear as pairs of EOFs, that could be degenerated representing similar patterns but shifted in phase (Horel 1984).
Figure~\ref{fig:eof-naive} shows the four eading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
It is clear that the first two EOFs represent the same zonal wave 1 pattern and the last two represent a same zonal wave pattern with shorter wavenumber and four centers of action shifted by 1/4 wavelength.
A similar EOF structure can be seen in 200 hPa (not shown).
Since each pair of EOFs seem to represent the same phase-varying structure, it would be desirable to combine them into a single index with amplitude and phase.

ON the other hand, complex Empirical Orthogonal Functions (cEOF) is a useful method to characterise zonal waves, associated with phase-varying structures (Horel 1984).
This method involves computing EOF on the analytic representation of the original field.
That representation is a complex field in which the real part is the original data and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
The Hilbert transform is usually understood in terms of time-varying signal.
However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.

@@

The phase of principal components is defined up to an additive constant.
For real principal components, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
For complex principal components, it can be any number between 0 and \(2\pi\) (Horel 1984).
Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real and imaginary parts are aligned with meaningful phases in our analysis.
This procedure does not create a spurious correlation, it only takes a known relationship and aligns it with a specific phase.
For the first complex principal component, the phase was chosen so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
This also nearly minimises the correlation with the imaginary part.
For the second complex principal component, the phase was chosen so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part was minimized, which also nearly maximizes the correlation with the imaginary part.

@@

\[
X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{Im}(\operatorname{cEOF}) + \beta(\lambda, \phi) \operatorname{Re}(\operatorname{cEOF}) + X_0(\lambda, \phi) +  \epsilon(\lambda, \phi, t)
\]

@@

We performed all analysis in this paper using the R programming language (R Core Team 2020), using data.table (Dowle and Srinivasan 2020) and metR (Campitelli 2020) packages. All graphics are made using ggplot2 (Wickham 2009). We downloaded data from reanalysis using the ecmwfr package (Hufkens 2020) and indices of the ENSO with the rsoi package (Albers and Campitelli 2020). The paper was rendered using knitr and rmarkdown (Y. Xie 2015; Allaire et al. 2020).

@@

\begin{figure*}
\includegraphics{../figures/extended-series-1} \caption{Time series extended using ERA5 back extended preliminary edition (period 1950 -- 1978) and ERA5 (period 1979 -- 2019). Dark straight line is the linear trend. Black horizontal and vertical line mark the mean value and range of each time series, respectively.}\label{fig:extended-series}
\end{figure*}

To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \ref{fig:ceofs-1} and \ref{fig:extended-series}.
The first mode (cEOF1) explains 82\% of the variance, while the second mode (cEOF2) explains a smaller fraction (7\%).
In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF.
The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}.

The cEOF1 (Fig.~\ref{fig:ceofs-1} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
The cEOF2 (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
This modulated amplitude is especially apparent at 200 hPa.
There is no apparent phase shift but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.

All cEOFs have non zero mean (Fig. \ref{fig:extended-series}) because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean.
However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.
There is no significant simultaneous correlation between the time series.
Both cEOFs show year-to-year variability but show no evidence of decadal variability.

A significant upward trend in the real component of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the complex components of cEOF2.
The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not in the phase (not shown).
This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
In agreement, M. Raphael (2003) detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.

\hypertarget{regressions}{%
\subsection{cEOFs Regressions}\label{regressions}}

@@

Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), both the Real and Imaginary cEOF1 are associated with planetary wave 1 patterns, that are 90º out of phase.
Their phases coincide with the ones shown in Figure~\ref{fig:ceofs-1}a.1, with the positive centre of the Real cEOF1 located towards the dateline, and the one of the Imaginary cEOF1 located over East Antarctica.
Instead of a wave-1 pattern, regressed anomalies describe a monopole with its centre displaced from the South Pole. The latter indicates an influence of cEOF1 in modulating the polar vortex.

The regression maps for cEOF1 at 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) are similarly influenced by the zonally symmetric circulation.
It is only possible to partially distinguish the wave 1 pattern in relation with the Real cEOF1 (Figure~\ref{fig:eof1-regr-gh}b.1).
The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase.

With the exception of the Imaginary cEOF1, it is clear that these patterns are very different than the fully zonally asymmetric versions (Fig.~\ref{fig:ceofs-1}), particularly at 200 hPa.
Moreover, only in the stratosphere these patterns actually show a distinguishable wave 1 pattern shifted in phase by 90º, suggesting that using the cEOF method is artificially generating a wave structure at 200hPa.
Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
While in the troposphere, they are associated with slightly off-centre monopoles.

\begin{figure}
\centering
\includegraphics{../figures/eof2-regr-gh-1.pdf}
\caption{\label{fig:eof2-regr-gh}Same as Figure~\ref{fig:eof1-regr-gh} but for the second cEOF.}
\end{figure}

Figure~\ref{fig:eof2-regr-gh} shows the regression pattern of geopotential height anomalies upon the cEOF2. Unlike for cEOF1, in this case the regression patterns are similar to the fully zonally asymmetric patterns from Figure~\ref{fig:ceofs-1}. Although there are some differences (particularly in 50 hPa), the wave trains identified before are well characterised and patterns associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2. Zonal wave 3 dominates all fields, but only in the western hemisphere, over the Pacific and Atlantic Oceans. cEOF2 then represents an equivalent barotropic wave train that is very similar to the the Pacific South American Patterns (Mo and Paegle 2001).
Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from Mo and Paegle (2001), the Real cEOF2 can roughly be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.

@@

\begin{figure}
\centering
\includegraphics{../figures/sam-eof-vertical-1.pdf}
\caption{\label{fig:sam-eof-vertical}Coefficient of determination between the real and imaginary part of each cEOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @Campitelli, Díaz, and Vera (2022) for the 1979 -- 2019 period. Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate (Benjamini and Hochberg 1995).}
\end{figure}

To explore the relationship between SAM and the modes described from the cEOF, we compute the coefficient of determination between the cEOFs timeseries and the three SAM indices (SAM, A-SAM and S-SAM) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
The SAM index is statistically signficant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
Correlations between SAM and the Real cEOF2 are non-significant.

The relationship between the tropospheric SAM and cEOF1 is explained entirely by the zonally symmetric component of the SAM as shown by the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1 and the high correlation with the S-SAM below 100 hPa.
In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
These correlations are consistent with the regression patterns of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by @Campitelli, Díaz, and Vera (2022).

In the case of cEOF2, the moderate correlation between SAM and Imaginary cEOF2 becomes extremely high only when the zonally asymmetric variability of the SAM is considered.
The Imaginary cEOF2 explains up to 92\% of A-SAM variance, reached the maximum value at 225 hPa (Figure~\ref{fig:sam-eof-vertical}.b2).
Such extremely high correlation between the asymmetric SAM and the Imaginary cEOF2 suggests that these might be different ways of characterising the same phenomenon.

@@

\caption{\label{tab:psa-eof2}Correlations between the Real and Imaginary parts of cEOF2 and the PSA1 and PSA2 modes computed as the second and third EOFs of seasonal geopotential height anomalies (following Mo and Paegle 2001) for the 1979 -- 2019 period. 95\% confidence intervals in parenthesis. p-values lower than 0.01 in bold.}
\centering
\begin{tabular}[t]{l>{}l>{}l}
\toprule
\multicolumn{1}{c}{} & \multicolumn{2}{c}{cEOF2} \\
\cmidrule(l{3pt}r{3pt}){2-3}
PC & Real & Imaginary\\
\midrule
PSA1 & 0.26 (CI: -0.04 -- 0.52) & \textbf{0.82 (CI: 0.69 -- 0.9)}\\
PSA2 & \textbf{0.79 (CI: 0.63 -- 0.88)} & -0.02 (CI: -0.32 -- 0.29)\\
\bottomrule
\end{tabular}
\end{table}

Due to the similarity between the PSA patterns and the cEOF2 spatial patterns (Fig. \ref{fig:eof2-regr-gh}), we study the relationship between these modes.
Table \ref{tab:psa-eof2} shows the correlations between the two PSA indices and the timeseries for Real and Imaginary phase of cEOF2.
As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong correlation between PSA1 and Imaginary cEOF2, and between PSA2 and Real cEOF2.
Conversely, there is no relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, as well as it is possible to make a rather clean association between its two phases and the two PSA modes.
So, It could be concluded that the same particular rotation of cEOF2 that maximises the association between cEOF2 parts and PSA modes, is the one which maximises the relationship between ENSO and Imaginary cEOF2.

The reason the conventional EOF analysis arrives at the same separation than our particular cEOF rotation is probably the fact that not all phases are equally frequent.
To visualize that, Figure~\ref{fig:phase-histogram} shows an histogram that counts the number of SON years in which the cEOF2 was close to each of the four particular phases (positive/negative Real/Imaginary), with the observations for each season marked as rugs on the horizontal axis.
For instance, years with cEOF2 phase within 45º of 0º are nearest the ``positive Real'' phase.
About two thirds of time cEOF2 has a phase similar to either the negative or positive Imaginary phase.
It could be seen that Imaginary phase is the most common phase and It is also, by construction, the direction that has the maximum relationship with ENSO.
Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.

\begin{figure}
\centering
\includegraphics{../figures/phase-histogram-1.pdf}
\caption{\label{fig:phase-histogram}Histogram of phase distribution of cEOF2 for the 1979 -- 2019 period. Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º. The small vertical lines near the horizontal axis mark the observations.}
\end{figure}

The result obtained here is similar to the one obtained by Irving and Simmonds (2016).
They show that the phase distribution of the PSA-like variability obtained by them is bimodal and isolate the PSA pattern from the rest of the PSA-like variability by selecting events that are near the peaks of the distribution (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).

The advantage of our method is that it is much simpler to implement, it provides magnitude and phase naturally, and it facilitates the description of this mode as a propagating wave instead of as standing oscillations.
As a consequence, the cEOF2 offers an alternative way of representing the PSA which has several advantages over using the second and third EOFs.

@@

>>>>

The Fourier decomposition relies on the assumption that the circulation can be meaningfully described in terms of zonal waves of constant amplitude along a latitude circle.
However, this is not valid for meridionally propagating waves or zonal waves with localised amplitudes.
Addressing this limitation, the Fourier technique can be generalized to integrate all planetary wave amplitude regardless of wave number, by computing the wave envelope (Irving and Simmonds 2015).
The latter makes it possible to represent planetary waves with different amplitude at different longitudes, but it removes all information about phase and wave number.
With this approach, Irving and Simmonds (2015) showed that planetary wave amplitude in general is associated to Antarctic sea-ice concentration and temperature, as well as to precipitation anomalies in regions of significant topography in SH mid-latitudes and Antarctica.

@@

An alternative methodology that has been proposed to study travelling and standing waves is complex Empirical Orthogonal Functions (Horel 1984).
This method extends EOF analysis to capture oscillations with varying amplitude and phase and has been applied to the time domain.
For instance, Krokhin and Luxemburg (2007) applied cEOF to station-based monthly precipitation anomalies and monthly temperature anomalies in the Eastern Siberia and the Far East region to characterise the main modes of variability and their connection to teleconnection indices.
Similarly, Gelbrecht, Boers, and Jürgen Kurths (2018) used cEOF applied to daily precipitation from reanalysis to study the propagating characteristics of the South American Monsoon.
To our knowledge, complex EOF has not been applied in the spatial domain to capture the phase-varying nature of planetary waves.

@@

In Section \ref{methods} we describe the methods.
In Section \ref{spatial} we analyse the spatial patterns of each complex EOF.
In Section \ref{regressions} we study the spatial regressions with geopotential height and temperature anomalies.
In Section \ref{other-variables} we analyse the relationship between cEOF1 and Ozone, and between cEOF2 with the SAM and the PSA.
In Section \ref{tropical} we study tropical forcings that explain the variability of each cEOF.
Finally, in Section \ref{precipitation} we show the relationship between these modes of variability and precipitation anomalies in South America and Oceania.

@@

We used monthly geopotential height, air temperature, ozone mixing ratio, and total ozone column (TOC) at 2.5\degree longitude by 2.5\degree latitude of horizontal resolution and 37 vertical isobaric levels from the European Centre for Medium-Range Weather Forecasts Reanalysis version 5 {[}ERA; Hersbach et al. (2019){]} for the period 1979 -- 2019.
Most of our analysis is restricted to the post-satellite era to avoid any confounding factors arising from the incorporation of satellite observations, but we also used the preliminary back extension of ERA5 from 1950 to 1978 (Bell et al. 2020) to describe long-term trends.
We derived streamfunction at 200 hPa from ERA5 vorticity using the FORTRAN subroutine FISHPACK (Adams, Swartztrauber, and Sweet 1999) and we computed horizontal wave activity fluxes following Plumb (1985).
Sea Surface Temperature (SST) monthly fields are from Extended Reconstructed Sea Surface Temperature (ERSST) v5 (Huang et al. 2017) and precipitation monthly data from the CPC Merged Analysis of Precipitation (P. Xie and Arkin 1997), with a 2.5\degree resolution in latitude and longitude.
The rainfall gridded dataset is based on information from different sources such as rain gauge observations, satellite inferred estimations and the NCEP-NCAR reanalysis, and it is available since 1979 to present.

The Oceanic Niño Index (ONI, Bamston, Chelliah, and Goldenberg 1997) comes from NOAA's Climate Prediction Center and the Dipole Mode Index (DMI, Saji and Yamagata 2003) from Global Climate Observing System Working Group on Surface Pressure.

@@

We computed the level-dependent SAM index as the leading EOF of year-round monthly geopotential height anomalies south of 20ºS at each level for the whole period (Baldwin and Thompson 2009).
We further split the SAM into its zonally symmetric and zonally asymmetric components (S-SAM and A-SAM indices respectively).
These indices were obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields, as proposed by Campitelli, Díaz, and Vera (2022).
Seasonal indices of the Pacific South American patterns (PSA1 and PSA2) were calculated, in agreement with Mo and Paegle (2001), as the third and fourth leading EOF of seasonal mean anomalies for 500-hPa geopotential heights at SH.

@@

In the standard EOF analysis, zonal waves may appear as pairs of (possibly degenerated) EOFs representing similar patterns but shifted in phase (Horel 1984).
Figure \ref{fig:eof-naive} shows the four leading EOFs of SON geopotential height zonal anomalies at 50 hPa south of 20\degree S.
It is clear that the first two EOFs represent a single phase-varying zonal wave 1 pattern and the last two represent a similarly phase-varying pattern with shorter wavenumber and four centres of action shifted by 1/4 wavelength.
A similar EOF structure can be seen in 200 hPa (not shown).
Since each pair of EOFs seems to represent the same phase-varying structure, it would be desirable to combine them into a single pattern described by amplitude and phase.

Complex Empirical Orthogonal Functions (cEOF) are a useful method for characterising zonal waves, associated with phase-varying structures (Horel 1984).
This method involves computing EOF on the analytic representation of the original field.
That representation is a complex field in which the real part is the original data and the imaginary part is the original data shifted by 90\degree at each spectral frequency -- i.e.~its Hilbert transform.
The Hilbert transform is usually understood in terms of time-varying signal.
However, in this work we apply the Hilbert transform at each latitude circle and at each considered time.
Since each latitude circle is a periodic domain, this procedure does not suffer from edge effects.

@@

The phase of cEOFs is defined up to an additive constant.
For real EOFs, this constant can be either 0 or \(\pi\), corresponding to a change in sign.
For cEOFs, it can be any real number between 0 and \(2\pi\) (Horel 1984), corresponding to rotations in the complex plane.
Since any choice is arbitrary and equally valid, we chose the phase of each cEOF so that the real or imaginary parts are aligned with meaningful phases in our analysis.

Preliminary analysis showed that the first cEOF was closely related to the the zonal wave 1 of Total Ozone Column and the second cEOF was closely related to ENSO.
Therefore, to aid in the interpretation, we chose the phases of each cEOF so that these relationships are aligned with one of the complex components.

We chose the phase of cEOF1 so that the time series corresponding to the real part has the maximum correlation with the zonal wave 1 of Total Ozone Column between 75°S and 45°S.
And we chose the phase of cEOF2 so that the coefficient of determination between the Oceanic Niño Index (Bamston, Chelliah, and Goldenberg 1997) and the real part is minimized, which also nearly maximizes the correlation with the imaginary part.

This procedure does not create spurious correlations, it only takes an existing relationship and aligns it with a specific phase.

@@

\[
X(\lambda, \phi, t) = \alpha(\lambda, \phi) \operatorname{Im}(\mathrm{cEOF}) + \beta(\lambda, \phi) \mathrm{Re}(\mathrm{cEOF}) + X_0(\lambda, \phi) + \epsilon(\lambda, \phi, t)
\]

@@

We performed all analysis in this paper using the R programming language (R Core Team 2020), using data.table (Dowle and Srinivasan 2020) and metR (Campitelli 2020) packages.
All graphics are made using ggplot2 (Wickham 2009).
We downloaded data from reanalysis using the ecmwfr package (Hufkens 2020) and indices of ENSO and Indian Ocean Dipole with the rsoi package (Albers and Campitelli 2020).
The paper was rendered using knitr and rmarkdown (Y. Xie 2015; Allaire et al. 2020).

@@

\begin{figure*}
\includegraphics{../figures/extended-series-1} \caption{Time series of the two leading cEOFs of SON zonal anomalies of geopotential height at 50 hPa and 200 hPa. cEOF1 (row a) and cEOF2 (row b) separated in their real (column 1) and imaginary (column 2) components. Dark straight line is the linear trend. Black horizontal and vertical line mark the mean value and range of each time series, respectively.}\label{fig:extended-series}
\end{figure*}

To describe the variability of the circulation zonal anomalies, the spatial and temporal parts of the first two leading cEOFs of zonal anomalies of geopotential height at 50 hPa and 200 hPa, computed jointly at both levels, are shown in Figures \ref{fig:ceofs-1} and \ref{fig:extended-series}.
The first mode (cEOF1) explains 82\% of the variance of the zonally anomalous fields, while the second mode (cEOF2) explains a smaller fraction (7\%).
In the spatial patterns (Fig.~\ref{fig:ceofs-1}), the real and the imaginary components are in quadrature by construction, so that each cEOF describe a single wave-like pattern whose amplitude and position (i.e.~phase) is controlled by the magnitude and phase of the temporal cEOF.
The wave patterns described by these cEOFs match the patterns seen in the traditional EOFs of Figure~\ref{fig:eof-naive}.

The cEOF1 (Fig.~\ref{fig:ceofs-1} column~a) is a hemispheric wave 1 pattern with maximum amplitude at high latitudes.
At 50 hPa the Real cEOF1 has the maximum of the wave 1 at 150ºE and at 200 hPa, the maximum is located at around 175ºE indicating a westerly shift in phase.
The cEOF2 (Fig.~\ref{fig:eof-naive} column~b) shows also a zonal wave-like structure with maximum amplitude at high latitudes, but with shorter spatial scales.
In particular, the dominant structure at both levels is a wave 3 but with larger amplitude in the pacific sector.
This modulated amplitude more evident at 200 hPa.
There is no apparent phase shift with height but the amplitude of the pattern is greatly reduced in the stratosphere, suggesting that this barotropic mode represents mainly tropospheric variability.

There is no significant simultaneous correlation between cEOFs time series.
Both cEOFs show year-to-year variability but show no evidence of decadal variability (Fig. \ref{fig:extended-series}).
Because the geopotential fields that enter into the cEOFs algorithm are anomalies with respect to the zonal mean instead of the time mean, the cEOFs have non zero temporal mean.
However, cEOF2's mean is almost zero, which indicates that only cEOF1 includes variability that significantly projects onto the mean zonally anomalous field.

A significant upward trend in the real component of cEOF1 is evident (Fig.~\ref{fig:extended-series}a.1, p-value~= 0.0037) while there is no significant trend in any of the complex components of cEOF2.
The positive trend in the Real cEOF1 translates into a positive trend in cEOF1 magnitude, but not systematic change in phase (not shown).
This long-term change indicates an increase in the magnitude of the high latitude zonal wave 1.
In agreement, M. Raphael (2003) detected a step change after around 1975 in the temporal evolution of leading EOF computed from August-September-October 500 hPa zonal geopotential height anomalies, which is similar to the cEOF1.

\hypertarget{regressions}{%
\subsection{cEOFs Regression maps}\label{regressions}}

@@

Figure~\ref{fig:eof1-regr-gh} shows regression maps of SON geopotential height anomalies upon cEOF1.
At 50 hPa (Figure~\ref{fig:eof1-regr-gh} row a), the Real cEOF1 is associated with a monopole centred over the Ross Sea.
The Imaginary cEOF1 is associated with a clear wave 1 pattern with maximum over the coast of East Antarctica.

At 200 hPa (Figure~\ref{fig:eof1-regr-gh} row b) the Real cEOF1 shows a single area of positive anomalies spanning West Antarctica surrounded by opposite anomalies in lower latitudes with is centre shifted slightly eastward compared with the upper level anomalies.
The Imaginary cEOF2.
The Imaginary cEOF1 shows a much more zonally symmetrical pattern resembling the negative SAM phase (Fogt and Marshall 2020).
Both components of cEOF1 could be associated to zonal anomalies in the upper troposphere jet stream intensity.

Therefore, the magnitude and phase of the cEOF1 are associated with the magnitude and phase of a zonal wave only in the stratosphere.
While in the troposphere, they are associated with slightly off-centre monopoles.

\begin{figure}
\centering
\includegraphics{../figures/eof2-regr-gh-1.pdf}
\caption{\label{fig:eof2-regr-gh}Same as Figure~\ref{fig:eof1-regr-gh} but for cEOF2.}
\end{figure}

Figure~\ref{fig:eof2-regr-gh} shows the regression maps of geopotential height anomalies upon the cEOF2.
Wave trains similar to those identified for cEOF2 patterns (Fig~\ref{fig:ceofs-1}) can be distinguished.
Regressed anomalies associated with the Real cEOF2 are 90º out of phase with those associated with the Imaginary cEOF2.
All fields have a dominant zonal wave 3 limited to the western hemisphere, over the Pacific and Atlantic Oceans.
cEOF2 then represents an equivalent barotropic wave train that is very similar to the the PSA Patterns (Mo and Paegle 2001).
Comparing the location of the positive anomaly near 90ºW in column b of Figure~\ref{fig:eof2-regr-gh} with Figures 1.a and b from Mo and Paegle (2001), the Real cEOF2 regression map could be identified with PSA2, while the Imaginary cEOF2 resembles PSA1.

@@

@@

\caption{\label{tab:psa-eof2}Correlation coefficients (r) between cEOF2 components and the PSA1 and PSA2 modes computed as Mo and Paegle (2001) for the 1979 -- 2019 period. 95\% confidence intervals in parenthesis. p-values lower than 0.01 in bold.}
\centering
\begin{tabular}[t]{l>{}l>{}l}
\toprule
\multicolumn{1}{c}{} & \multicolumn{2}{c}{cEOF2} \\
\cmidrule(l{3pt}r{3pt}){2-3}
PC & Real & Imaginary\\
\midrule
PSA1 & 0.26 (CI: -0.04 -- 0.52) & \textbf{0.82 (CI: 0.69 -- 0.9)}\\
PSA2 & \textbf{0.79 (CI: 0.63 -- 0.88)} & -0.02 (CI: -0.32 -- 0.29)\\
\bottomrule
\end{tabular}
\end{table}

Given the similarity between the cEOF2 regression maps (Fig.~\ref{fig:eof2-regr-gh}) and documented PSAs we study the relationship between these modes.
Table~\ref{tab:psa-eof2} shows the correlations between the two PSA indices and the time series for Real and Imaginary phases of cEOF2.
As anticipated by Figure~\ref{fig:eof2-regr-gh}, there is a strong positive correlation between PSA1 and Imaginary cEOF2, and also between PSA2 and Real cEOF2.
On the other hand, there is no significant relationship between PSA1 and Real cEOF2, and between PSA2 and Imaginary cEOF2.
As a result, cEOF2 represents well both the spatial structure and temporal evolution of the PSA modes, so it is possible to make an association between its two phases and the two PSA modes.
The same phase election for cEOF2 that maximises the relationship between ENSO and Imaginary cEOF2, also maximises the association between cEOF2 components and PSA modes.

\begin{figure}
\centering
\includegraphics{../figures/phase-histogram-1.pdf}
\caption{\label{fig:phase-histogram}Histogram of phase distribution of cEOF2 phase for the 1979 -- 2019 period. Bins are centred at 90º, 0º, -90º, -180º with a binwidth of 90º. The small vertical lines near the horizontal axis mark the observations.}
\end{figure}

Figure~\ref{fig:phase-histogram} shows an histogram that counts the number of SON seasons in which the cEOF2 phase was close to each of the four particular phases (positive/negative of real/imaginary component), with the observations for each season marked as rugs on the horizontal axis.
In 62\% of seasons cEOF2 has a phase similar to either the negative or positive imaginary phase, making the imaginary phase the most common phase.
This is also, by construction, the phase that is most correlated with ENSO.
Therefore, the Imaginary cEOF2 explains more variance than the Real cEOF2 and conventional EOF analysis will tend to separate the two.

This phase preferences is in agreement with Irving and Simmonds (2016), who found a bimodal distribution to PSA-like variability (compare our Figure~\ref{fig:phase-histogram} with their Figure 6).

\begin{figure}
\centering
\includegraphics{../figures/sam-eof-vertical-1.pdf}
\caption{\label{fig:sam-eof-vertical}Coefficient of determination (\(r^2\)) between each component of cEOFs and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level for the 1979 -- 2019 period. Thick lines represent estimates with p-value \textless{} 0.01 corrected for False Detection Rate (Benjamini and Hochberg 1995).}
\end{figure}

We now explore the relationship between SAM and the cEOFs motivated by the resemblance between cEOFs regression maps and SAM patterns shown in the previous section.
To that end, we computed the coefficient of determination between the cEOFs time series and the three SAM indices (SAM, A-SAM and S-SAM) defined by Campitelli, Díaz, and Vera (2022) at each vertical level (Fig.~\ref{fig:sam-eof-vertical}).
The SAM index is statistically significant correlated with the Real cEOF1 in all levels, and with the Imaginary cEOF1 and Imaginary cEOF2 in the troposphere.
On the other hand, correlations between SAM and the Real cEOF2 are non-significant.

The relationship between the SAM and cEOF1 in the troposphere is explained entirely by the zonally symmetric component of the SAM as shown by the high correlation with the S-SAM below 100 hPa and the low and statistically non-significant correlations between the A-SAM and either the Real or Imaginary cEOF1.
In the stratosphere, the Real cEOF1 is correlated with both A-SAM and S-SAM, while the Imaginary cEOF1 is highly correlated only with the A-SAM.
These correlations are consistent with the regression maps of geopotential height in Figure~\ref{fig:eof1-regr-gh} and their comparison with those obtained for SAM, A-SAM and S-SAM by Campitelli, Díaz, and Vera (2022).

In the case of Imaginary cEOF2, the correlation between SAM and Imaginary cEOF2 for the troposphere is associated to the asymmetric variability of the SAM.
Indeed, the Imaginary cEOF2 shares up to 92\% variance with the A-SAM and only 12\% at most with the S-SAM (Figure~\ref{fig:sam-eof-vertical}.b2).
Such extremely high correlation between A-SAM and Imaginary cEOF2 suggests that the methodology proposed in this work is able to characterise the zonally asymmetric component of the SAM described previously by Campitelli, Díaz, and Vera (2022).

@@

\begin{itemize}
\item
  The Oceanic Niño Index is available via NOAA's Climate Prediction Center: \url{https://www.cpc.ncep.noaa.gov/products/analysis_monitoring/ensostuff/detrend.nino34.ascii.txt}.
\item
  The Dipole Mode Index is available via Global Climate Observing System Working Group on Surface Pressure: \url{https://psl.noaa.gov/gcos_wgsp/Timeseries/Data/dmi.had.long.data}
\end{itemize}
